---
title: "class1"
author: "Sandrine Mfuranzima"
date: "2025-10-16"
output: html_document
---
### Data structure (Vectors, Factors, Lists, Data frame, matrix and arrays)

#### Vector: The foundational R data structure
##### Numeric
```{r}
#example
x <- c(0.5, 0.6)
age <- c(20, 35, 32, 29)
summary(age)

# Reassigning the age with new values
age <- c(35, 24, 18, 24)

# Mean of new age values
mean(age)

```
###### Logical
```{r}
x <- c(TRUE,FALSE)
x <- c(T,F)
```
###### Character

```{r}
x <- c("a","b","c")
class <-c("M","F","F","M")
```
###### Integer

```{r}
x <-9:29
```

###### Complex

```{r}

x<- c(1+0i, 2+4i)
x
```

###### Matrix and Operation within Matrices

```{r}
x<-matrix(1:6, nrow=2 ,ncol=3)
x

y<-matrix(1:6,nrow =3 ,ncol =2 )
y

x%*%y
```

###### List

```{r}
x<-list(1,"a",TRUE,1+4i)
x
```

###### Factor

```{r}
x<-factor(c("yes","yes","no","yes","no"))
x
```

###### Create a vector with NAs in it

```{r}
x<-c(1,2,NA,10,3)
##check which number in  X is NA 
is.na(x)
```

```{r}
x <- c(1, 2, 4, "NA", 5)
bad <- is.na(x)
print(bad)
x[!bad]
```

```{r}
x <- c(1, 2, NA, 4, NA, 5)
y <- c("a", "b", NA, "d", NA, "f")
good <- complete.cases(x, y)
good
```
### Coercion
```{r}
sum(c(TRUE, TRUE, FALSE, FALSE, FALSE))
```
```{r}
# Let's Create a Data Frame
studentID <- c(1,2,3,4,5)
math_score <- c(12,17,10,9,NA)
gender <- c("M","F","M","M","F")
it_score <- c(13,18,11,10,19)
scoredata <- data.frame(studentID, gender, math_score, it_score)
scoredata
#View(scoredata)
```

##### Create data from keyboard
```{r}
data_class2<-data.frame(height=numeric(0),weight=numeric(0),bmi=numeric(0))
data_class2<-edit(data_class2)
```

##### Data Importation

```{r}
data_class<-read.csv("C:\\Users\\DXB\\Desktop\\R_programing\\class3\\random_height_weight_dataset.csv")
variable.names(data_class)
View(data_class)
```

```{r}
head(data_class)
```

```{r}
tail(data_class)
```
```{r}
data_class[10:20,]
```
```{r}
summary(data_class$W)
```
```{r}
length(data_class$v2)
```
```{r}
data_class[,-1]
```


```{r}
#Built-in Dataset Example
data("airquality")
variable.names(airquality)
str(airquality)
summary(airquality)
```
### Exploration Data Analysis (EDA)

```{r}
data_class <- data.frame(
HEIGHT = c(161,155,158,170,160,156,162,158,158,167,160,155,154,155,157,157,160,158,160,160,
152,154,150,161,162,164,161,155,159,163,159,160,158,165,156,163,155,164,163,168,
157,164),
WEIGHT = c(50,49,42,65,60,52,58,46,45,51,60,42,53,48,48,48,53,52,51,53,
44,56,63,52,57,49,52,54,46,50,61,55,45,63,60,56,52,47,52,55,
48,58)
)
summary(data_class)
apply(data_class,2,mean)
apply(data_class,2,sd)
```
### Relationship Between Variables – iris Dataset
```{r}
data("iris")
head(iris)
mean(iris$Petal.Length)
mean(iris$Petal.Length[iris$Species=="setosa"])
mean(iris$Petal.Length[iris$Species=="versicolor"])
mean(iris$Petal.Length[iris$Species=="virginica"])

#shortcut
by(iris$Petal.Length,iris$Species,mean)
by(iris$Petal.Length,iris$Species,sd)
by(iris$Petal.Length,iris$Species,summary)

```
##### Visualize the differences in continuous variables between categories using Box-and- whisker plot

```{r}
boxplot(iris$Sepal.Length ~ iris$Species, data = iris, main="Comparison")
```


### Relationship Between Two Categorical Variables
```{r}
ucba<-data.frame(UCBAdmissions)
head(ucba)
cross<-xtabs(Freq~Gender+Admit,data=ucba)
(cross <- xtabs(Freq ~ Gender + Admit, data = ucba))
prop.table(cross, 2)
```
### Simpson's paradox
```{r}
cross2<-xtabs(Freq~Gender+Admit,data=ucba[ucba$Dept=="A",])
prop.table(cross2,1)
```

##### Bar Chart
```{r}
dat <- read.table(text ="ProdA ProdB ProdC ProdD
1 110 50 60 70
2 120 50 80 65", header=TRUE)
barplot(as.matrix(dat), beside=FALSE, col=c("red","green"))


dat1 <- read.table(text = "A B C D E F G
1 10 80 30 90 70 60 90
2 20 50 70 50 40 10 40
3 60 80 80 60 60 30 160
4 20 40 70 80 20 10 70", header = TRUE)
barplot(as.matrix(dat1))
```
### Scatter Plot
```{r}
plot(WEIGHT ~ HEIGHT, data=data_class)
plot(data_class$HEIGHT,data_class$WEIGHT)
plot(data_class[,1],data_class[,2])
cor(data_class$HEIGHT, data_class$WEIGHT)
```

### Linear Model

```{r}
plot(WEIGHT~HEIGHT,data=data_class) 
abline(lm(WEIGHT~HEIGHT,data=data_class)$coefficient)
```
```{r}
data_lm<-lm(WEIGHT~HEIGHT,data=data_class) 
summary(data_lm)
hist(data_class$HEIGHT)
```
```{r}
hist(data_class$WEIGHT,col = "blue")
```

```{r}
plot(WEIGHT~HEIGHT,data=data_class) 
abline(lm(WEIGHT~HEIGHT,data=data_class)$coefficient) 
lines(lowess(data_class$HEIGHT,data_class$WEIGHT),col="blue")
```
```{r}
cor(data_class$HEIGHT,data_class$WEIGHT) 
model1<-lm(WEIGHT~HEIGHT,data=data_class)
```
### Data Management – Creating BMI
```{r}
BMI <- round(data_class$WEIGHT / (data_class$HEIGHT/100)^2, 2)
new_data_class <- cbind(data_class, BMI)
head(new_data_class, n=5)
summary(BMI)
stem(BMI, scale=2)
```


